The Talking Heads experiment

### 2015-04-29 chap1/chap1.tex
part 1 of martin haspelmath's comments

The Talking Heads experiment is first and foremost
a scientific experiment. It subjects four radical ideas 
to experimental scrutiny. 
The first idea is that 
language emerges through self-organisation out of local
interactions of language users. It spontaneously 
becomes more complex to increase
reliability and optimise transmission across generations of 
users, without a central designer.
I call this {\itshape the selfish
language hypothesis}: Language colonises
brains and recruits available cognitive capacities to satisfy
its appetite for expressing ever more complex meaning with
minimal effort and maximum 
effectiveness.\footnote{
See \cite{Deacon:1998} for a discussion of 
co-evolution between cognitive and linguistic
capacities and brain structures.}
Language is 
not a uniform abstract system of rules (and definitely not an 
innate system of rules) but a creative open-ended complex
adaptive system, like a natural
ecology, in which certain solutions to relate forms with 
meanings become temporarily conventionalised in 
the community, even though new creative solutions
emerge almost any time someone speaks. Language
is in constant flux because new meanings continuously
arise and existing forms undergo change beyond the 
control of any individual language user.

The second radical idea is that 
meaning is built up slowly by each individual in a cumulative
growth process. Meaning is not innate, as rationalists in the 
footprints of Plato have been arguing for centuries, nor 
learned through stepwise
induction from examples and counterexamples, as empiricists
have been saying. Meaning is at first very concrete and
strongly situated in the environment and bodily experiences.
I will take the suggestions made
by Wittgenstein \is{Wittgenstein} one step further, namely that meaning (and language) is constructed
and practised as part of
language games.\footnote{\cite{Wittgenstein:1953} emphasises the relativity 
of concepts and the role of language and hence
meaning in social interactions.}
I will introduce a selectionist approach to
the acquisition of meaning, introducing models to 
show that conceptual distinctions can \smplenquote{grow} in the brain
like the leaves and branches on a tree and be pruned to fit the
demands and characteristics of the environment in which
an agent finds itself. Even though non-verbal 
activities, like predicting the future based on a model 
or deciding what to do in specific circumstances, stimulates
the growth of distinctions, language use is probably one 
of the greatest stimulators of conceptual growth. It provides
feedback about which distinctions were successful in linguistic
communication and thus whether or not they should be preserved. 
Thus language and cognition co-evolve. Each one pushes the other 
up towards more complexity and they become tightly 
co-ordinated with neither a central co-ordinator nor prior innate design.

@@

In 1996, a computer program called Deep Blue
defeated the reigning world champion Kasparov in a game of 
chess \citep{Newborn:1996}.
Kasparov was astonished
and depressed, and claimed this was a defeat of humans in 
the 
race against machines. But was he actually beaten by {\itshape artificial}
intelligence? Not really. A team of engineers and scientists
from Carnegie Mellon University and from the IBM Watson 
Research Center had been working for ten years to program 
vast amounts of chess knowledge, invented by
human experts, into Deep Blue. They had built extremely 
sophisticated dedicated computing hardware to
apply this knowledge at a blinding speed. So, in beating Kasparov, 
other humans were the clever ones, not machines.

@@

The Talking Heads experiment is the culmination of one of
the most exciting scientific and engineering projects I have
ever been involved in. It has required the
creative efforts of a dozen excellent researchers 
over many years. The story started in 
1985. Instead of continuing to design and program intelligence
explicitly based on formalising human cognitive
capacities, as most of my colleagues
in artificial intelligence research 
labs were doing,\footnote{
Some recent overviews of this \smplenquote{classical} approach to 
artificial intelligence can be found in: \cite{Nilsson:1998} and 
\cite{Russell:1998}.}
I started to focus on the 
question of how intelligence might
originate and evolve in physical agents as they 
interact autonomously with their environment or with 
other humans, and I encouraged my students
at the Artificial Intelligence Laboratory of the University 
of Brussels (VUB) to experiment in the same direction. 
We initially developed a bottom-up, 
behaviour-oriented approach to sensori-motor intelligence, 
which was also being explored around the same time by 
Rodney Brooks at the MIT Artificial Intelligence 
Laboratory.
The behaviour-oriented, bottom-up approach \is{behaviour-based robots} was a 
counter reaction to the symbolic, top-down approach of 
earlier AI research. See \cite{Steels:1995} and \cite{Arkin:1998}. 
We built robots of various sizes and shapes, using simple
electronic circuits, Lego bricks, small motors, rechargeable
batteries, self-made 
sensors, single board computers, and everything else that 
appeared useful (\figref{f:plate3}).

@@

In the summer of 1995, a clear breakthrough occurred.  
I was working as a visiting researcher in the Sony
Computer Science Laboratory in Tokyo, invited by its director
Mario Tokoro. Reflecting on our 
experiments from a distance, two new ideas occurred to me. 
First of all, language may have been the missing key
in the initial experiments. Language may 
be a necessary route by which the human cognitive system
bootstraps itself autonomously, in tight interaction with the
environment and aided by a community of other
language speakers. This suggested that if we wanted
to have emergent
forms of cognitive intelligence, we needed to go the same 
route. Second, the principles and mechanisms that had been
pouring out of the study of complexity had to be
relevant to understanding the origins and evolution of language,
because they provided generic explanations for how complexity 
may emerge. These principles include 
self-organisation, structural coupling, 
selectionism, level formation, 
and many others (\citealt{Nicolis:1989}). The field of {\itshape artificial life} brings together researchers
exploring the insights of complex systems with computational
and robotic experiments (see: \citealt{Langton:1995}).

Together with 
other researchers interested in the then-arising field of \smplenquote{artificial life}, I had 
already been simulating path formation in 
ant societies and other biological phenomena exhibiting 
an emergence of complexity. It dawned on me that the
importance of these mechanisms for bootstrapping intelligence and
language might be much greater than 
thought so far.\footnote{The following reference provides a general survey of similar work in 
the area of lexicon formation: \cite{Steels:97b}. 
A representative sample of work on syntax is \cite{Briscoe:1999}.}

\clearpage
Back in Brussels, an extremely exciting but very 
intense period of research started as I tried to 
apply these principles to language
and subject them to experimental 
scrutiny.\footnote{The earliest papers on these mechanisms are in: \cite{Steels:95b} and 
\cite{Steels:96a}.}
We very quickly built a first 
prototype of a {\itshape Talking Head} with our own hardware
and low-level software (see \figref{f:plate7}) 
and experimented with language games on mobile 
robotic agents.\footnote{The electronics and tracking software for this 
active camera were built by Tony Belpaeme. The mobile
robot experiments were conducted with Paul Vogt. 
Other early work on the grounding and autonomous acquisition of 
language-like communication systems by robotic systems is 
described in \cite{Steels:97g}. See also: \cite{Billard:1998}.}
New fundamental insights and discoveries emerged almost daily. 
The first experiments, particularly in grounding
language on real robots, were extremely difficult but we 
were clearly making steady progress.

@@

\enlargethispage{1\baselineskip}
The massive response and thoughtful judgements of the 
public were crucial to validate many aspects of the 
theories put forward in this book. 
But the interactions with a broad public and
the intense discussions that it generated also added new
dimensions to the research. First of all, a whole 
new type of interface between man and machine was
taking shape under our very eyes. In contrast to pre-programmed
computer interfaces, which more often than not make it
difficult to do what one wants, the Talking Heads demonstrated
for the first time the concept of {\itshape negotiated user
interfaces}. The interaction was based on mutual respect
and adaptation of man {\itshape and} machine. Communicative
failure was not fatal but 
an opportunity to fine-tune and negotiate the way communication
would take place in the future.

>>>>

The Talking Heads experiment is first and foremost
a scientific experiment. It subjects four radical ideas 
to experimental scrutiny. 
The first idea is that 
language emerges through self-organisation out of local
interactions of language users. It spontaneously 
becomes more complex to increase
reliability and optimise transmission across generations of 
users, without a central designer.
I call this {\scshape the selfish
language hypothesis}: Language colonises
brains and recruits available cognitive capacities to satisfy
its appetite for expressing ever more complex meaning with
minimal effort and maximum 
effectiveness.\footnote{
See \cite{Deacon:1998} for a discussion of 
co-evolution between cognitive and linguistic
capacities and brain structures.}
Language is 
not a uniform abstract system of rules (and definitely not an 
innate system of rules) but a creative open-ended complex
adaptive system, like a natural
ecology, in which certain solutions to relate forms with 
meanings become temporarily conventionalised in 
the community, even though new creative solutions
emerge almost any time someone speaks. Language
is in constant flux because new meanings continuously
arise and existing forms undergo change beyond the 
control of any individual language user.

The second radical idea is that 
meaning is built up slowly by each individual in a cumulative
growth process. Meaning is not innate, as rationalists in the 
footprints of Plato have been arguing for centuries, nor 
learned through stepwise
induction from examples and counterexamples, as empiricists
have been saying. Meaning is at first very concrete and
strongly situated in the environment and bodily experiences.
I will take the suggestions made
by Wittgenstein \is{Wittgenstein} one step further, namely that meaning (and language) is constructed
and practised as part of
language games.\footnote{\cite{Wittgenstein:1953} emphasises the relativity 
of concepts and the role of language and hence
meaning in social interactions.}
I will introduce a selectionist approach to
the acquisition of meaning, introducing models to 
show that conceptual distinctions can \dblenquote{grow} in the brain
like the leaves and branches on a tree and be pruned to fit the
demands and characteristics of the environment in which
an agent finds itself. Even though non-verbal 
activities, like predicting the future based on a model 
or deciding what to do in specific circumstances, stimulates
the growth of distinctions, language use is probably one 
of the greatest stimulators of conceptual growth. It provides
feedback about which distinctions were successful in linguistic
communication and thus whether or not they should be preserved. 
Thus language and cognition co-evolve. Each one pushes the other 
up towards more complexity and they become tightly 
co-ordinated with neither a central co-ordinator nor prior innate design.

@@

In 1996, a computer program called Deep Blue
defeated the reigning world champion Kasparov in a game of 
chess \citep{Newborn:1996}.
Kasparov was astonished
and depressed, and claimed this was a defeat of humans in 
the 
race against machines. But was he actually beaten by {\scshape artificial}
intelligence? Not really. A team of engineers and scientists
from Carnegie Mellon University and from the IBM Watson 
Research Center had been working for ten years to program 
vast amounts of chess knowledge, invented by
human experts, into Deep Blue. They had built extremely 
sophisticated dedicated computing hardware to
apply this knowledge at a blinding speed. So, in beating Kasparov, 
other humans were the clever ones, not machines.

@@

The Talking Heads experiment is the culmination of one of
the most exciting scientific and engineering projects I have
ever been involved in. It has required the
creative efforts of a dozen excellent researchers 
over many years. The story started in 
1985. Instead of continuing to design and program intelligence
explicitly based on formalising human cognitive
capacities, as most of my colleagues
in artificial intelligence research 
labs were doing,\footnote{
Some recent overviews of this \dblenquote{classical} approach to 
artificial intelligence can be found in: \cite{Nilsson:1998} and 
\cite{Russell:1998}.}
I started to focus on the 
question of how intelligence might
originate and evolve in physical agents as they 
interact autonomously with their environment or with 
other humans, and I encouraged my students
at the Artificial Intelligence Laboratory of the University 
of Brussels (VUB) to experiment in the same direction. 
We initially developed a bottom-up, 
behaviour-oriented approach to sensori-motor intelligence, 
which was also being explored around the same time by 
Rodney Brooks at the MIT Artificial Intelligence 
Laboratory.
The behaviour-oriented, bottom-up approach \is{behaviour-based robots} was a 
counter reaction to the symbolic, top-down approach of 
earlier AI research. See \cite{Steels:1995} and \cite{Arkin:1998}. 
We built robots of various sizes and shapes, using simple
electronic circuits, Lego bricks, small motors, rechargeable
batteries, self-made 
sensors, single board computers, and everything else that 
appeared useful (\figref{f:plate3}).

@@

In the summer of 1995, a clear breakthrough occurred.  
I was working as a visiting researcher in the Sony
Computer Science Laboratory in Tokyo, invited by its director
Mario Tokoro. Reflecting on our 
experiments from a distance, two new ideas occurred to me. 
First of all, language may have been the missing key
in the initial experiments. Language may 
be a necessary route by which the human cognitive system
bootstraps itself autonomously, in tight interaction with the
environment and aided by a community of other
language speakers. This suggested that if we wanted
to have emergent
forms of cognitive intelligence, we needed to go the same 
route. Second, the principles and mechanisms that had been
pouring out of the study of complexity had to be
relevant to understanding the origins and evolution of language,
because they provided generic explanations for how complexity 
may emerge. These principles include 
self-organisation, structural coupling, 
selectionism, level formation, 
and many others (\citealt{Nicolis:1989}). The field of {\scshape artificial life} brings together researchers
exploring the insights of complex systems with computational
and robotic experiments (see: \citealt{Langton:1995}).

Together with 
other researchers interested in the then-arising field of \dblenquote{artificial life}, I had 
already been simulating path formation in 
ant societies and other biological phenomena exhibiting 
an emergence of complexity. It dawned on me that the
importance of these mechanisms for bootstrapping intelligence and
language might be much greater than 
thought so far.\footnote{The following reference provides a general survey of similar work in 
the area of lexicon formation: \cite{Steels:97b}. 
A representative sample of work on syntax is \cite{Briscoe:1999}.}

\clearpage
Back in Brussels, an extremely exciting but very 
intense period of research started as I tried to 
apply these principles to language
and subject them to experimental 
scrutiny.\footnote{The earliest papers on these mechanisms are in: \cite{Steels:95b} and 
\cite{Steels:96a}.}
We very quickly built a first 
prototype of a {\scshape Talking Head} with our own hardware
and low-level software (see \figref{f:plate7}) 
and experimented with language games on mobile 
robotic agents.\footnote{The electronics and tracking software for this 
active camera were built by Tony Belpaeme. The mobile
robot experiments were conducted with Paul Vogt. 
Other early work on the grounding and autonomous acquisition of 
language-like communication systems by robotic systems is 
described in \cite{Steels:97g}. See also: \cite{Billard:1998}.}
New fundamental insights and discoveries emerged almost daily. 
The first experiments, particularly in grounding
language on real robots, were extremely difficult but we 
were clearly making steady progress.

@@

\enlargethispage{1\baselineskip}
The massive response and thoughtful judgements of the 
public were crucial to validate many aspects of the 
theories put forward in this book. 
But the interactions with a broad public and
the intense discussions that it generated also added new
dimensions to the research. First of all, a whole 
new type of interface between man and machine was
taking shape under our very eyes. In contrast to pre-programmed
computer interfaces, which more often than not make it
difficult to do what one wants, the Talking Heads demonstrated
for the first time the concept of {\scshape negotiated user
interfaces}. The interaction was based on mutual respect
and adaptation of man {\itshape and} machine. Communicative
failure was not fatal but 
an opportunity to fine-tune and negotiate the way communication
would take place in the future.
